A Bayesian geostatistical approach to modelling     
global distributions of *Lygodium microphyllum*     
under projected climate warming     
==================================================
     
Authors:
--------
John M. Humphreys (jmh09r@my.fsu.edu)    
James B. Elsner   
Thomas H. Jagger  
Stephanie Pau   
     
    
The below script and linked data are sufficient to specify and fit the full model from the above cited paper.  Be advised, run time for the model may take several hours.     
    
An extended and executed version of this script (with all models/figures) can be viewed here:    
http://rpubs.com/JMHumphreys/LygPA020217              
   
  
```{r}
date()
```

####Options   
```{r}
library(knitr)
library(rgl)
opts_knit$set(verbose = FALSE)
knit_hooks$set(webgl = hook_webgl)
```
      
###Load Packages:
```{r}
suppressMessages(library(INLA))
#suppressMessages(library(rgdal))
#suppressMessages(library(reshape2))
suppressMessages(library(raster))
#suppressMessages(library(fields))
suppressMessages(library(rasterVis))
suppressMessages(library(rgeos))
suppressMessages(library(Thermimage))
suppressMessages(library(maptools))
suppressMessages(library(mapproj))
#suppressMessages(library(spdep))
suppressMessages(library(ggplot2))
#suppressMessages(library(GISTools))
suppressMessages(library(lattice))
suppressMessages(library(gridExtra))
suppressMessages(library(dplyr))
```


###Get Project Data
The link provided below will download all of the data needed to run this script.  The text file includes the coordinates for species occurrences and qudrature as well as all covariates.  Descriptions of the data fields are provided below.  
    
     
Data and copies of the R-Markdown files are available at:https://github.com/JMHumphreys/SDM_Lygodium    

To view extended versions of the executed code, visit:  https://rpubs.com/JMHumphreys/LygPA020217     
```{r}
LygPA.df = read.csv("https://github.com/JMHumphreys/SDM_Lygodium/blob/master/LygodiumData.csv?raw=true",
             stringsAsFactors = FALSE)

dim(LygPA.df)
head(LygPA.df)
```
    
###Dataframe to spatialpoints   
The resulting spatialpointsdataframe will be projected to 3D Cartesian coordinates below.  
```{r}
LL84 = "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"

LL = cbind(LygPA.df$Long, LygPA.df$Lat)
LygPA = SpatialPointsDataFrame(LL, LygPA.df)
proj4string(LygPA) = LL84
```
   
   
###Get Country Boundaries     
```{r}
Countries = map("world", 
                fill = TRUE, 
                plot = FALSE)


IDs = sapply(strsplit(Countries$names, ":"), function(x) x[1])

LL84 = "+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0"

CountriesP = map2SpatialPolygons(Countries, IDs = IDs,
                                proj4string = CRS(projection(LL84 )))

#Add a dataframe      
pid = sapply(slot(CountriesP, "polygons"), 
             function(x) slot(x, "ID"))

p.df = data.frame( ID=1:length(CountriesP), 
                   row.names = pid)

Countries = SpatialPolygonsDataFrame(CountriesP, p.df)


#Rasterized version for later plotting     
Ras = raster(res = 0.9,
             crs = proj4string(Countries))

Domain.r = rasterize(Countries, Ras, 
                     field = 0, 
                     background = NA)

```

###Locations of Observations in 2D   
```{r}
LygC = subset(LygPA, OBS == 1) #Species observations

Cp = levelplot(Domain.r, 
               margin = FALSE,
               xlab = NULL, ylab = NULL, 
               col.regions = "tan",
               colorkey = NULL, 
                     panel = panel.levelplot, space = "right", 
                     par.strip.text = list(fontface='bold'),
                     par.settings = list(axis.line = list(col = "black"), 
                                              strip.background = list(col = 'transparent'), 
                                              strip.border = list(col = 'transparent')), 
                                              scales = list(col = "black", cex = 1, relation="free")) + 
              latticeExtra::layer(sp.polygons(Countries, col = "black", lwd = 1)) +
              latticeExtra::layer(sp.polygons(LygC, col = "red", pch=19, cex=0.5))

Cp
```
    
    
###Construct a mesh for the global domain ("mesh1")   
The mesh is constructed to have a greater number of triangulation over terrestrial regions.   
```{r warning = FALSE, webgl=TRUE}
CountriesU = gBuffer(Countries, width = 1) #Smoothing edges

MaxEdge = 2.0

bdry = inla.sp2segment(CountriesU)

mesh2D = inla.mesh.2d(boundary = bdry, 
                     cutoff = 1, 
                     max.edge = MaxEdge,
                     min.angle = 21) 

MeshLocs = cbind(mesh2D$loc[,1], mesh2D$loc[,2]) #Locations of mesh nodes

#Converting to 3D coordinates
xyz = as.data.frame(                
          inla.mesh.map(MeshLocs, 
                        projection = "longlat", 
                        inverse = TRUE))

true.radius.of.earth = 6371
radius.of.earth = 1

mesh1 = inla.mesh.create(loc = xyz,
                         cutoff = 400/true.radius.of.earth, 
                         refine=list(max.edge = 3500/true.radius.of.earth, min.angle = 26))

mesh1$n
plot(mesh1, rgl = TRUE, main = " ")
```

   
   
###Convert to Cartesian Coordinates
Specify the "spde" spatial effect.  Converting the 2D long-lat coordinates from the observation and quadrature locations to 3D Cartesian coordinates (xyz) and then associating them to the mesh.  Also creating a spatial index ("field0").              
```{r}
spde = inla.spde2.pcmatern(mesh1,
                           prior.range=c(0.9, 0.9),  
                           prior.sigma=c(1, 0.01))

locs = cbind(LygPA$Long, LygPA$Lat)

locs = inla.mesh.map(locs, 
                     projection = "longlat", 
                     inverse = TRUE)


A = inla.spde.make.A(mesh1, loc=locs) #project point locations to mesh

field0 = inla.spde.make.index("field0", 
                                  spde$n.spde) #spatial index
```
     

###Construct Data Stack  
Data stacks are used to put all fixed and random variables into a more easily managed object.  Although not used in this script, the variables "mTcm70" and "Pwq70" included with "LygPA" dataset contain projected temperature and precipitation data for the year 2070 as described in the publication.      
```{r}
FE.df = LygPA@data


FE.lst = list(c(field0,                     #Spatial Index
                list(intercept0 = 1)),      #Intercept
                list(mTcm = FE.df[,"mTcm"], #Min Temp of Coldest Month
                     Pwq = FE.df[,"Pwq"],   #Precipitation of Wettest Quarter
                     Pop = FE.df[,"Pop"],   #Human Population density
                     CTI = FE.df[,"CTI"],   #Compound Topographic Index
                     NN = FE.df[,"NN"],     #Distance to nearest plant occurrence
                     CoH = FE.df[,"CoH"]))  #Invasive Cohort Richness
                  

#Stack0 for model fitting
Stack0 = inla.stack(data = list(OBS = FE.df$OBS), #Response.  Presence (1) or Quadrature (0)  
                                  A = list(A, 1), #Intersection of points with mesh    
                            effects = FE.lst,     #Index, Intercept, and Covariates  
                                tag = "obs0")     #Just a label 
```

           
###Modify SPDE Model  
Before adding covariates to the model, the "spde" specification is modified.  The "extraconstr" control directs the SPDE spatial effect to a "design matrix" that holds the covariates and intercept; this helps reduce confounding by ensuring that the spatial effect is orthogonal to the covariates.  A "PC" Prior is specified based on the mesh parameters.   
```{r}
n.data = dim(LygPA@data)[1]
n.covariates = 7

X = cbind(rep(1,n.data), 
        FE.df$mTcm,
        FE.df$Pwq,
        FE.df$Pop,
        FE.df$CTI,
        FE.df$NN,
        FE.df$CoH)

Q = qr.Q(qr(X))


spde = inla.spde2.pcmatern(mesh1,
                           prior.range=c(0.9, 0.9),
                           prior.sigma=c(1, 0.01),  
                           extraconstr = list(A = as.matrix(t(Q)%*%A), 
                                              e = rep(0, n.covariates)))  
```
   
###Full Model (Model5)
```{r}
pcprior1 = list(prec = list(prior="pc.prec", param = c(3, 0.01))) 
pcprior2 = list(prec = list(prior="pc.prec", param = c(0.1, 0.01)))

#theta5 = Model5$internal.summary.hyperpar$mean
theta5 = c(-0.89178665, -0.06363225, -0.32293059, -0.64617359) #thetas from prior run

Frm5 = OBS ~ -1 + intercept0 + 
                  f(field0, model=spde) +
                  f(mTcm, model = "rw1",  
                    hyper = pcprior1,
                    scale.model = TRUE) + 
                  f(NN, model = "rw1",  
                    hyper = pcprior2,
                    scale.model = TRUE) + 
                 Pop + CTI + Pwq + CoH          


Model5 = inla(Frm5, 
               data = inla.stack.data(Stack0, spde=spde), 
               family = "binomial", 
               verbose = TRUE,
               control.fixed = list(prec = 0.1, prec.intercept = 0.1),    
               control.predictor = list(
                                      A = inla.stack.A(Stack0), 
                                      compute = TRUE, 
                                      link = 1), 
               control.mode = list(restart = TRUE, theta = theta5),
               control.results = list(return.marginals.random = TRUE,
                                      return.marginals.predictor = TRUE),
               control.compute=list(dic = TRUE, cpo = TRUE, waic = TRUE)) 
```
```{r}
summary(Model5) 
```
   
             
###Random Effects  
```{r}
proj = inla.mesh.projector(mesh1, 
                           dims=c(400, 800))

plotdata = inla.mesh.project(proj, Model5$summary.random$field0[,"mean"])

R600 = raster(nrows = 400, ncols = 800)

values(R600) = plotdata

M600 = as.matrix(R600)

M0.spm = rotate90.matrix(M600) #rotation before rasterization
M0.spmr = raster(M0.spm)
extent(M0.spmr) = c(-180, 180, -90, 90)
proj4string(M0.spmr) = CRS(proj4string(Domain.r))


rng = seq(-4, 13, 0.01)
cr = colorRampPalette(c("darkblue", "blue", "lightblue",
                        "yellow", "orange", "red", "darkred"),  
                        bias = 1.5, space = "rgb")

Cp = levelplot(M0.spmr, 
               margin = FALSE,
               xlab = NULL, ylab = NULL, 
               col.regions = cr, at = rng,
               colorkey = list(at=seq(-5, 13, 0.01), 
                     labels=list(at=c(-4, 0, 4, 8, 12)), 
                     labels=c("-4", "0", "4", "8", "12")), 
                     panel = panel.levelplot, space = "right", par.strip.text = list(fontface='bold'),
                     col = cr, par.settings = list(axis.line = list(col = "black"), 
                                              strip.background = list(col = 'transparent'), 
                                              strip.border = list(col = 'transparent')), 
                                              scales = list(col = "black", cex = 1))
Cp + 
  latticeExtra::layer(sp.polygons(Countries, col = "black", lwd = 1))
```


###Standard Deviation    
```{r}
plotdata = inla.mesh.project(proj, Model5$summary.random$field0[,"sd"])

R600 = raster(nrows = 400, ncols = 800)

values(R600) = plotdata

M600 = as.matrix(R600)

M0.spm = rotate90.matrix(M600) #rotation before rasterization
M0.spmr = raster(M0.spm)
extent(M0.spmr) = c(-180, 180, -90, 90)
proj4string(M0.spmr) = CRS(proj4string(Domain.r))

rng = seq(0, 4.4, 0.01)
cr = colorRampPalette(c("lightblue", "yellow", "darkred"),  
                        space = "rgb")

Cp = levelplot(M0.spmr, 
               margin = FALSE,
               xlab = NULL, ylab = NULL, 
               col.regions = cr, at = rng,
               colorkey = list(at=seq(0, 4.4, 0.01), 
                     labels=list(at=c(0, 1, 2, 3, 4)), 
                     labels=c("0", "1", "2", "3", "4")), 
                     panel = panel.levelplot, space = "right", par.strip.text = list(fontface='bold'),
                     col = cr, par.settings = list(axis.line = list(col = "black"), 
                                              strip.background = list(col = 'transparent'), 
                                              strip.border = list(col = 'transparent')), 
                                              scales = list(col = "black", cex = 1))
Cp + 
  latticeExtra::layer(sp.polygons(Countries, col = "black", lwd = 1))
```


###Temperature Limitation   
Reduced survival in cold temperatures.     
```{r}
mic.df = as.data.frame(Model5$summary.random$mTcm[,1:6])
names(mic.df) = c("ID", "Mean", "sd", "Q025", "Q50", "Q975")

ggplot(mic.df, aes(ID*10, Mean)) +
        geom_smooth(col = "black", 
                  linetype= "solid",
                  method = "loess",
                  span = 0.25,
                  se = FALSE,
                  lwd = 1) +
        geom_smooth(data = mic.df, aes(ID*10, Q025), 
                    col = "grey", 
                    method = "loess",
                    span = 0.25,
                    se = FALSE,
                    linetype= "dashed") +
        geom_smooth(data = mic.df, aes(ID*10, Q975), 
                    col = "grey", 
                    method = "loess",
                    span = 0.25,
                    se = FALSE,
                    linetype= "dashed") +
        geom_hline(yintercept = 0, 
                   linetype = "dotted",
                   col = "red",
                   size = 1) +  
        geom_vline(xintercept = 0, 
                   linetype = "dotted",
                   col = "red",
                   size = 1) +
        xlab(expression("Limiting Temperature  ("*~degree*C*"  )")) +
        ylab("Logit") +  
        theme_classic() +
        theme(axis.text=element_text(size=16),
              axis.title.y = element_text(size = 20),
              axis.title.x = element_text(size = 20, vjust=-2))
```

###Find Limiting Temperture    
Point at which logit becomes positive.    
```{r}
FindZero = function(X.df) {

      Cross = which(diff(sign(X.df$Mean))!=0)

      mTemp = (X.df$ID[Cross] + X.df$ID[Cross+1])/2

      return(mTemp)      
}

###Temperature Limit (degrees C) #3.85
FindZero(mic.df)[2]*10

```

###Fine-scale Spatial Structure     
Accounting for the observed clustering pattern (possible related to dispersal ability).    
```{r}
mic.df = as.data.frame(Model5$summary.random$NN[,1:6])
names(mic.df) = c("ID", "Mean", "sd", "Q025", "Q50", "Q975")

Full.p = ggplot(mic.df, aes(ID*1000, Mean)) +
        geom_smooth(method = "loess",
                    se = FALSE, col = "black", 
                    linetype= "solid") +
        geom_smooth(data = mic.df, aes(ID*1000, Q025), 
                    method = "loess",
                    se = FALSE, col = "grey", 
                    linetype= "dashed") +
        geom_smooth(data = mic.df, aes(ID*1000, Q975), 
                    method = "loess",
                    se = FALSE, col = "grey", 
                    linetype= "dashed") +
        geom_hline(yintercept = 0, 
                   linetype = "dotted",
                   col = "red",
                   size = 1) + 
        geom_vline(xintercept = 0, 
                   linetype = "dotted",
                   col = "red",
                   size = 1) +
        xlim(c(0,28000)) +
        xlab(NULL) +
        ylab("Logit") +
        ggtitle("A") +
        theme_classic() +
        theme(axis.text=element_text(size=16),
              axis.title.y = element_text(size = 20),
              axis.title.x = element_text(size = 20, vjust=-2),
              plot.title = element_text(hjust = 0.5))



###Zoom in some
mic.df = as.data.frame(Model5$summary.random$NN[,1:6])
names(mic.df) = c("ID", "Mean", "sd", "Q025", "Q50", "Q975")

mic.df = mic.df %>%
          filter(ID <= 2)

Zoom.p = ggplot(mic.df, aes(ID*1000, Mean)) +
        geom_smooth(method = "loess",
                    se = FALSE, col = "black", 
                    linetype= "solid") +
        geom_smooth(data = mic.df, aes(ID*1000, Q025), 
                    method = "loess",
                    se = FALSE, col = "grey", 
                    linetype= "dashed") +
        geom_smooth(data = mic.df, aes(ID*1000, Q975), 
                    method = "loess",
                    se = FALSE, col = "grey", 
                    linetype= "dashed") +
        geom_hline(yintercept = 0, 
                   linetype = "dotted",
                   col = "red",
                   size = 1) + 
        geom_vline(xintercept = 0, 
                   linetype = "dotted",
                   col = "red",
                   size = 1) +
        xlim(c(0,2000)) +
        xlab("Distance to Nearest Neighbor (km)") +
        ylab("Logit") +  
        ggtitle("B") +
        theme_classic() +
        theme(axis.text=element_text(size=16),
              axis.title.y = element_text(size = 20),
              axis.title.x = element_text(size = 20, vjust=-2),
              plot.title = element_text(hjust = 0.5))


grid.arrange(Full.p, Zoom.p, ncol=1)
```

###Threshold Distance
```{r}
###Max Distance for dispersal #1585
FindZero(mic.df)[1]*1000
```

